---
title: "Exploratory Analysis"
author: "Nathan Bush"
date: "2022-11-08"
output:
  html_document: default
  pdf_document: default
---
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(tigris)
library(raster)
library(readxl)
library(sf)
library(tmap)
library(equatiomatic)
library(gridExtra)
```


```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
load("~/Documents/GitHub/Data-Analytics-Term-Project/Relevant Files/dataset_final_new.Rdata")
load("~/Documents/GitHub/Data-Analytics-Term-Project/Relevant Files/color.Rdata")
load("~/Documents/GitHub/Data-Analytics-Term-Project/Relevant Files/states.Rdata")
load("~/Documents/GitHub/Data-Analytics-Term-Project/Relevant Files/final_map.Rdata")
load("~/Documents/GitHub/Data-Analytics-Term-Project/Relevant Files/final_dataset.Rdata")
load("~/Documents/GitHub/Data-Analytics-Term-Project/Relevant Files/final_dataset.Rdata")

```

REPORT
  
In this analysis I compare the political leanings of states in the US with the their energy usage statistics and emissions. This is important to understand whether or not political views effect the type of energy states are using. This is important because is can effect and inform policy, show whether or not states are taking advantage of their natural resources, and help usunderstand the dynamics behind energy systems and how politics plays a role in these systems. If we find that conservative leaning states tend to use less renewable energy than their state's natural resources provide, or that liberal states tend to rely excessively on renewable energy when their state is lacking in these resources, then we can assume that more effective policy needs to be implemented to incentivize states to act more efficiently with their energy generation.
  
The data I use in this analysis come from three sources. My first dataset is energy usage statistics on a wide variety of metrics in each state. This is taken from the US Energy Information Administration. Key variables here are the percentages of petroleum, natural gas, coal, and renewable energy as a percentage of the total energy use in the state. My second and third datasets are county voting data on the 2020 presidential election taken from the MIT Election and Data Science lab and the Harvard Dataverse. The variables here are State, County, Candidate, # of votes in each county, total votes. My third dataset is table that webscraped from Wikipedia showing who each state's governor is and which political party they are associated with. My hypothesis is that a state who voted for a republican in the 2020 election and has a republican governor is more conservative than a state that has a democratic governor but voted republican in the 2020 election. I will be able to test whether or not there is any difference in energy usage based on these observed differences by adding in the governor variable. There is a fourth dataset in the start of my code but that did not end up being used.

Cleaning and merging this data was a timely process. I cleaned the energy data by getting rid of unnecessary rows and columns generated in excel and deleting variables unrelated to my analysis. Some of the data had * for values less than 0.5 which were deemed as statistically equal to 0. I changed these values to 0 and adjusted some names to be easier to work with. I deleted total US observations and the DC observation in addition to transposing key rows and columns to get the tidy in tidy format. I made a few other adjustments that are notated in my code as well. For the presidential data, I filtered out all observations for votes other than those for Joe Biden and Donald Trump. I also needed to summarize all vote types, create a new variable that summarized total votes in across counties within states, and create a dummy variable for party. The governor's data was fairly simple to web scrape and import into r. I deleted a a couple of unnecessary variables and named key variables accordingly. 

I merged the data using state as the unique ID and then got to work creating 3 state political categories. Blue, Red, and Purple, I used dummy variables for governor (1 if republican, 0 if not), for president (1 if republican, 0 if not). I then created a variable for state political category where (Blue = 0 for governor dummy & 0 for president dummy, Red = 1 for governor dummy and 1 for president dummy, Purple = 1 in one dummy, 0 in another). I then created another variable mean renewable energy percentage based on state political category and showed this graphically. The units of analysis in this final merged data set vary between variables and are specific to each respective visualization.

I did employ a few transformations throughout the course the project. These are outlined in each visualization, but are primarily based upon adjusting state energy measurement statistics to account for differing state GDP's and populations. I did not remove any outliers in the raw data.


Visualization 1
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
graph_1_color <- ggplot(data=color, aes(x=state_cat, y=renew,fill=state_cat))+
  geom_bar(stat="identity")+
  scale_fill_manual(values=c("blue",
                             "purple",
                             "red")) + labs(y= "% Renewable Energy" , x = "State Category") + 
  theme(legend.position = "none") + geom_col(colour = "black") +  geom_label(aes(label = renew), vjust = 1.5, colour = "black", fill = "white") +
  ylim(0,33) +labs(title = "Electricity Generation by State Political Category")
graph_1_color
```
Based on the above graph, we can see the percentage of electricity among groups of states that is generated by renewable energy. We observe that purple states (states with governors whose political party does not align with the way the state voted in the 2020 election) actually produce less of their electricity using renewable energy than either blue states (states with a democratic governor who voted for Joe Biden in 2020) or red states (states with a republican governor who voted for Donald Trump in 2020). This is an interesting result. One hypothesis for this could be that purple states are less efficient at passing legislation due to their conflicting political idealogies than either blue or red states. If renewable energy is objectively in the best interest of agents, passing renewable energy legislation would only be held up by one political attempting to spite the other for leverage purposes.


Visualization 2
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
affiliation_map <- tm_shape(final_map)+tm_polygons(col ="state_category", palette = c(Blue = "blue", Red = "red", Purple = "purple"), title = "Political Category") 
affiliation_map

```
This map is showing us the state category of each state in the US by color. We can use this as a reference for our results going forward. I removed Alaska and Hawaii from this map for visual reasons. Alaska is red, Hawaii is blue. 


Visualization 3
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
energy_map = tm_shape(final_map)+tm_polygons(col ="Renewables", palette =c("Greens"), title = "Renewable Energy (%)")
energy_map
```
This map shows the renewable energy percentage by state. Dark states have higher renewable energy percentages. We can see that states west of the Mississippi and concentrated in the northwest and northeast corners of the country seem to have higher percentages of renewable energy production. Again, Alaska and Hawaii have been removed for visual reasons here. Alaska has 43% of its electricity generated by renewable energy, and Hawaii has 19%. 



Visualization 4
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
dot_plot =  ggplot(dataset_final_new, aes(x = Renewables, y = reorder (state, Renewables), col = state_category, label = state)) +
    geom_point() + scale_color_manual(values=c("blue", "purple", "red")) +
  ylab("State") + xlab("Renewable Energy (% Electricity)") +
  theme(legend.position = "none")

dot_plot_2 = ggplot(dataset_final_new, aes(x = Renewables, y = reorder (state, Renewables))) +
  geom_point(size = 3, aes(colour = state_category)) +
  scale_color_manual(values=c("blue", "purple", "red"), limits = c("Blue","Purple","Red")) +
  facet_grid(state_category ~ ., scales = "free_y", space = "free_y") +
  theme(legend.position = "none") +  
  ylab("State") + xlab("Renewable Energy (% Electricity)")

grid.arrange(dot_plot, dot_plot_2, ncol=2)
```
In the above graph, I show the percentage of renewable energy by state in two dot plots. The plot on the left shows Renewable Energy percentages across all 50 states. We can see that the amount of renewable energy each states uses spans a the entire range between 0% and 100%, although it seems appears that roughly 2/3 of states use less than 25% renewable energy. On the right had side, I show the same graph, but separated by state political category. We do appear not see any noteworthy trends based upon political category. Although Vermont, a purple state, appears to be a bit of an outlier in the purple category, and is ran on 100% renewable energy. Very cool.

Visualization 5
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
graph_5.0 = ggplot(data = final_dataset, aes(x = new_solar, y = avg_temp, col = state_category, label = state, group = state_category)) +
  geom_point() + scale_color_manual(values=c("blue", "purple", "red")) + 
  geom_text(hjust=1, vjust=1.5, size = 3) + 
  ylab("Average Temperature") + 
  xlab("Solar Energy") + geom_smooth(method="lm", se = FALSE)
graph_5.0
```

In this a graph I show the relationship between solar energy and average temperature by state. Intuition would lead us to believe that states with higher average temperatures employ more solar energy because they likely have more sunlight. I add local regressions to my graph to test whether or not this is the case. I find evidence that blue and red states are indeed employing more solar energy in states with hotter temperatures, while purple states are showing the opposite trend.Perhaps blue and red states are acting rationally in their own best interest, while purple states are unable to do so because of an entangled political environment. This MAY lend credence to our initial hypothesis that states with conflicting political interests are less efficient than states with more uniform political environments. 

Visualization 6
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
graph_6 = ggplot(data = final_dataset, aes(x = Renewables, y = co2_emissions, col = state_category, label = state, group = state_category)) +
  geom_point() + scale_color_manual(values=c("blue", "purple", "red")) + 
  geom_text(hjust=1, vjust=1.5, size = 3) + 
  ylab("CO2 Emissions") + 
  xlab("Renewable Energy %") 
graph_6
```
In this graph I am looking to see if there is a relationship between Renewable Energy used and CO2 emissions. I am interested to see if there is correlation between the amount of Renewable energy used and amount of CO2 emitted.Based on this graph, we see that Texas is by far the biggest Co2 emitter in the US, despite using 25% renewable energy. Next, I will divide CO2 emissions by population size to see CO2 emissions per capita.

Visualization 7
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
final_dataset = final_dataset |> mutate(co2_pop = co2_emissions/labor_force)

graph_6.0 = ggplot(data = final_dataset, aes(x = Renewables, y = co2_pop, col = state_category, label = state, group = state_category)) +
  geom_point() + scale_color_manual(values=c("blue", "purple", "red")) + 
  geom_text(hjust=1, vjust=1.5, size = 3) + 
  ylab("CO2 Emissions") + 
  xlab("Renewable Energy %")

graph_6.0
```

Now I get a graph showing CO2 emissions per capita and Renewable energy %. We see that Wyoming, West Virginia, and South Dakota are clear outliers and are significantly above the national average of CO2 emissions per capita. It is worth noting that all three of these states are red. Texas has entered the middle of the pack now that we adjust for state population. Let's remove these three outliers and add local regressions by state political category to observe any potential relationship.

Discussion of ideal data...consumption and production and differences between the 2
difference between consumption and generation




Visualization 8
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
graph_6.0.1 = ggplot(data = final_dataset, aes(x = Renewables, y = co2_pop, col = state_category, label = state, group = state_category)) +
  geom_point() + scale_color_manual(values=c("blue", "purple", "red")) + 
  geom_text(hjust=1, vjust=1.5, size = 3) + 
  ylab("CO2 Emissions") + 
  xlab("Renewable Energy %") +ylim(0,25000) +
  geom_smooth(method = "lm", se = FALSE)

graph_6.0.1
```

Finally, we see that there is a negative linear relationship between CO2 emissions per capita and the percentage of renewable energy each state uses. This relationship is similar within each state political category. One observation worth noting is that of the 20 states with CO2 emissions per capita of less than 7500 metric tons, just 3 of these states are "red". 

I included the three graphs above to show my data visualization process and why it is useful to visualize your data with graphs. In visualization 6, we clearly saw that Texas was an outlier, calling for a transformation of our CO2 emissions variable to account for state population. Then, in visualization 7, we clearly see that there are three outlier states which will severely impact our analysis unless removed. Finally, we arrive at visualization 8 which gives us a more complete and holistic picture of the true relationship between renewable energy % and CO2 emissions. 

REGRESSIONS
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align='center'}
final_dataset$GDP =  as.numeric(final_dataset$GDP)
final_dataset$avg_temp =  as.numeric(final_dataset$avg_temp)
final_dataset$Renewables =  as.numeric(final_dataset$Renewables)
final_dataset$precip =  as.numeric(final_dataset$precip)
```

```{r, results='hide'}
reg_1 = feols(Renewables ~ avg_temp + GDP + precip, data = final_dataset)
summary(reg_1)
```

```{r, results='hide'}
reg_2 = feols(Renewables ~ avg_temp + GDP + precip + co2_pop + trump_vote * rep_gov, data = final_dataset)
summary(reg_2)
```

```{r, results='hide'}
reg_3 = feols(Solar ~ GDP + precip + Renewables + state_category + avg_temp, data = final_dataset)
summary(reg_3)
```

```{r}
etable(reg_1,reg_2, reg_3)
```
#MOTIVATION#


In my econometric analysis, I will be using regression analysis as a descriptive tool. I am interested in examining the relationship between a couple of independent variables within my data and my key variable of interest, the amount of renewable energy that each state generates as a percentage of its total electricity consumption. These relationships are important to analyze to see if there sort of predictive power or relationship between the variables. 


#METHODS AND RESULTS#


My first regression equation is: 

latex
$$Renewable Energy = \beta_{0} + \beta_{1} Average Temperature + \beta_{2} Precipitation + \beta_{3} GDP + \beta_{4} Purple + \beta_{5} Red $$
Based on the table above, we see that both average temperature and precipitation have statistically significant negative coefficients at the 10% level. A 1 unit increase in the average temperature corresponds with a -1.138 change in the percent of renewable energy a state uses. A 1 unit increase in the amount of precipitation a state recieves corresponds to a -0.5504 percent change in the amount of renewable energy a state uses. This is likely due to a lack of sunshine, and the may mean that solar energy is a relatively more popular form of renewable energy generation, an interesting result. The effect of GDP is statistically 0. 


My 2nd Regression is: 

latex
$$Renewable Energy = \beta_{0} + \beta_{1} Average Temperature + \beta_{2} Precipitation + \beta_{3} GDP + \beta_{4} Purple + \beta_{5} Red + \beta_{6} CO2 Emissions + \beta_{7} Trump Vote + \beta_{8} Republican Governor + \beta_{9} Trump Vote X Republican Governor$$
In this equation, I add a continuous variables for CO2 emissions per capita, and fully interact binary variables for a Trump Vote and Republican Governor. In this model, we see that a state that both voted for Trump, and has a republican governor, is predicted to use 5.874% less renewable energy in comparison to states with just a trump vote or a republican governor, or neither. however, this result is not statistically significant. We also observe that a 1 unit increase in CO2 emissions per capita is associated with a .0004 percent decline in renewable energy, significant at the 10% level. 

My final regression is 

$$Solar Energy = \beta_{0} + \beta_{1} Average Temperature + \beta_{2} GDP + \beta_{3} Precipitation + \beta_{4} Renewables + \beta_{5} State Category + \beta_{6} CO2 Emissions + \beta_{7} Trump Vote + \beta_{8} Republican Governor + \beta_{9} Trump Vote X Republican Governor$$



In the above section, I estimate 4 regression equations. 
