---
title: "State Politics & Energy Usage"
author: "Nathan Bush"
date: Advanced Data Analytics
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
load("~/Documents/GitHub/Data Analytics Term Project/Relevant Files/graph_1_color.Rdata")
load("~/Documents/GitHub/Data Analytics Term Project/Relevant Files/energy_map.Rdata")
load("~/Documents/GitHub/Data Analytics Term Project/Relevant Files/affiliation_map.Rdata")
load("~/Documents/GitHub/Data Analytics Term Project/Relevant Files/graph_6.Rdata")
load("~/Documents/GitHub/Data Analytics Term Project/Relevant Files/graph_6.0.Rdata")
load("~/Documents/GitHub/Data Analytics Term Project/Relevant Files/graph_6.0.1.Rdata")
load("~/Documents/GitHub/Data Analytics Term Project/Relevant Files/dot_plot.Rdata")
load("~/Documents/GitHub/Data Analytics Term Project/Relevant Files/dot_plot_2.Rdata")
load("~/Documents/GitHub/Data Analytics Term Project/Relevant Files/graph_5.0.Rdata")
library(gridExtra)
```

## Introduction

1. **Potential Topics**
    - Fish populations in the Salmon river 
    - National Park visitation and Social Media
 <br>

2. **Motivation**
    - Curious about state politics & how it effects decision making
    - Energy is a good metric to measure this 
    - Interesting independently 
    <br>
    
3. **Hypothesis**
    - Conservative states use less energy than is optimal
    
    
---

## Why is this important?

  - Do state politics interfere with decision making?
  
    
    
   - Are states acting in their own best interest?
   
  
    
  - Understand these dynamics to best inform policy
    
---

## Data

**Energy usage statistics by state**
       - US Energy Information Administration
        
**2000-2020 presidential election data**
        -MIT Election and Data Science Lab & the Harvard Dataverse 
        
**2022 State Governors** 
        -Webscraped from Wikipedia
---

# Data Cleaning Process

**Energy Use Data**
- This dataset needed the most work
- Removed rows containing non-data entries
- Removed variables unnecessary to my analysis
- Applied transformations to key variables 
    - (* = 0 | --, NM = N/A)

**Election Data**
- Clean Dataset...tricky variable creation
- Summed votes for candidates within states to determine victor
- Dummy variable = 1 if a state voted for Trump

**Governor Data**
- Webscraped Wikipedia Table
- Easy to clean and process
- Dummy variable = 1 if a state has a republican governor

---

**Question 1**

Are there discernible differences between renewable energy use based on political affiliation between states?

Three Categories
- Red
- Blue
- Purple

---
```{r fig.align='center'}
affiliation_map
```
---
```{r fig.align='center'}
energy_map
```
---
```{r fig.align='center'}
graph_1_color
```





